---
title: "Powstream Gap Analysis"
author: "Lindsay Carr, Alison Appling, Jordan Read"
date: "November 17, 2015"
---

```{r echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, eval=TRUE, results='hide'}

library(unitted)
library(ggplot2)
library(knitr)
library(dplyr)
library(tidyr)

allData_list <- load.gapanalysis.data()

for(i in 1:length(allData_list)){
  eval(parse(text=paste0(names(allData_list)[i], " <- allData_list[[", i, "]]")))
}

```

This analysis will compare conditions at stream sites that have metabolism estimates with all NHDPlus stream sites in the US. The goal is to identify which conditions of US streams have not been fully represented by the collection of current metabolism sites. Identification of such underrepresented conditions will help in choosing 30 new monitoring sites to create a more robust dataset for metabolism estimates. 

Data for NHDPlus streams comes from Mike Wieczorek's *NHDPlus Version 2 Accumulated and Reach Catchment Variables* dataset on ScienceBase. Data representing the current metabolism sites comes from using the `get_meta()` function in `powstreams.R` and has data for the 291 sites with metabolism estimates. Seven metrics have been used in this analysis; additionally, each metric has an *Accumulated Watershed* (watershed down the NHDPlus network at each catchment) density plot and a *Reach Catchment* plot (watershed for each specific NHDPlus catchment).  

  
**Metrics used for this analysis:**

* Stream Order: metabolism sites have larger streams than is seen nationally in NHD+
* Watershed Area: metabolism sites have slightly large watershed areas than all NHD+ reaches
* Runoff: metabolism sites do not capture the extreme low flows that are seen in NHD+ reaches
* Precipitation: metabolism sites cover the range of precipitation, except for the extreme lows
* Temperature: metabolism sites cover the range of temperatures, except for the extreme lows
* Agriculture: metabolism sites represent NHD+ reaches agricultural activity well, but miss the extreme highs
* Population Density: metabolism sites represent higher populated areas than seen nationally
* Ecoregions: metabolism sites represent eastern ecoregions well, but miss much of the western ecoregions
* State Discharge & DO: maps showing locations of the metabolism sites in relation to existing discharge measurement sites


```{r echo=FALSE, warning=FALSE, message=FALSE}
allsites_data <- list(allData_land, allData_land_RE, allData_pop, allData_precip, allData_run, allData_temp)
num_allsites <- max(sapply(allsites_data, function(x) {length(unique(x$COMID))}))
num_metabsites <- length(unique(metabData$site_num))
num_sites_table <- data.frame(Source = c("Metabolism", "All US"), 
                              `Number of Sites`=c(num_metabsites, num_allsites))
kable(num_sites_table)
```

\newpage




##Stream Order

Stream order is a representation of the size and complexity of a stream or river. Stream order values used in this comparison come from NHDPlus, which uses the Strahler stream ordering methodology to calculate stream order values. 

```{r order, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}

metab_order <- filter(allData_order, comid %in% metabData$nhdplus_id)
df_order <- format.df("streamorde", "streamorde", metab_order, allData_order, type=NULL) 

plot.orderHist(allData_order, df_order)

```

\newpage


##Watershed Area

The drainage area for either the accumulated watershed or the reach catchment, measured in square meters.

```{r watershed, echo=FALSE, warning=FALSE, message=FALSE}

df_area_AC <- format.df("landcover.ACCUM_AREA", "ACLU_AREA", metabData, allData_land, type=NULL) 
df_area_RE <- format.df("landcover.REACH_AREA", "NLCD11_ARE", metabData, allData_land_RE, type=NULL)

df_area <- format.catchment.df(df_area_AC, df_area_RE)
plot.dens(df_area, "Watershed Areas", log=TRUE, reach=TRUE, type="", xlabel="Watershed Area (sq. meters)")

```

\newpage

```{r runoff, echo=FALSE}
```

##Runoff

Runoff data is the average estimated runoff based on streamflow records from 1971-2000, and is measured in millimeters (mm). 

Note: no estimation method was reported with this dataset. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

df_run_AC <- format.df("hydrol.RUNOFF_AC", "RUNOFF_AC", metabData, allData_run, type=NULL)
df_run_RE <- format.df("hydrol.RUNOFF_RE", "MEAN", metabData, allData_run, type=NULL)

df_run <- format.catchment.df(df_run_AC, df_run_RE) 
plot.dens(df_run, "Average Runoff\n(1971-2000)", log=TRUE, reach=TRUE, type="runoff",
          xlabel="Runoff (mm)")

```

Note: The presence of zero values in the dataset skews this density plot because 0s are treated as NAs on a log scale. Additionally, there values below 0.1 mm are not shown on this density plot because there were so few observations. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
kable(format.runoff.count(df_run))
```

\newpage

```{r precip, echo=FALSE}
```

##Precipitation

Precipitation data is the average annual precipitation (cm/year) from 1971-2000 based on 800 meter PRISM data. A correction has been applied to the data coming from Science Base for NHD+ Sites at the reach catchment scale. For the metabolism sites, Ted Stets divided the reach catchment values by 1000; the same correction has been applied for reach-scale precipitation values coming from the NHD+ dataset.

```{r echo=FALSE, warning=FALSE, message=FALSE}

df_precip_AC <- format.df("climate.PPT30YR_AC", "PPT30YR_AC", metabData, allData_precip, type=NULL)
df_precip_RE <- format.df("climate.PPT30YR_RE", "MEAN", metabData, allData_precip, type="re_precip")

df_precip <- format.catchment.df(df_precip_AC, df_precip_RE)

#convert from cm to mm
df_precip$data_vals <- df_precip$data_vals*10

plot.dens(df_precip, "Average Annual Precipitation\n(1971-2000)", log=FALSE, reach=TRUE, type="",
          xlabel="Precipitation (mm)")

```


\newpage


```{r temp, echo=FALSE}
```

##Temperature

Temperature data is the average annual air temperature (in deg C) from 1971-2000 based on 800 meter PRISM source grids.    

```{r echo=FALSE, warning=FALSE, message=FALSE}

df_temp_AC <- format.df("climate.TMEAN_AC", "TMEAN_AC", metabData, allData_temp, type=NULL)
df_temp_RE <- format.df("climate.TMEAN_RE", "MEAN", metabData, allData_temp, type=NULL)

df_temp <- format.catchment.df(df_temp_AC, df_temp_RE)
plot.dens(df_temp, "Average Temperature\n(1971-2000)", log=FALSE, reach=TRUE, type="",
          xlabel="Temperature (deg C)")

```


\newpage



```{r agriculture, echo=FALSE}
```

##Agriculture

Data representing agriculture is the sum of the land-use classes 81 (Pasture/Hay) and 82 (Cultivated Crops). The values are shown as the percent of the watershed area covered in pasture/hay or cultivated crops.    

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=5}

df_ag_crop_AC <- format.df("landcover.AC_NLCD11Buffpct82", "ACLU11_82", metabData, allData_land, type="ac_ag")
df_ag_hay_AC <- format.df("landcover.AC_NLCD11Buffpct81", "ACLU11_81", metabData, allData_land, type="ac_ag")
df_ag_AC <- data.frame(data_type = df_ag_crop_AC$data_type, 
                       data_vals = df_ag_crop_AC$data_vals + df_ag_hay_AC$data_vals)

df_ag_crop_RE <- format.df("landcover.RE_NLCD11Buffpct82", "NLCD11_82", metabData, allData_land_RE, type="re_ag")
df_ag_hay_RE <- format.df("landcover.RE_NLCD11Buffpct81", "NLCD11_81", metabData, allData_land_RE, type="re_ag")
df_ag_RE <- data.frame(data_type = df_ag_crop_RE$data_type, 
                       data_vals = df_ag_crop_RE$data_vals + df_ag_hay_RE$data_vals)

df_ag <- format.catchment.df(df_ag_AC, df_ag_RE) 
mainplot <- plot.dens(df_ag, "Fractional Coverage of Agriculture\n(National Land Cover Database 2011)", 
                      log=FALSE, reach=TRUE, type="", xlabel="Fraction of Pasture + Crops in Watershed Area")

rect_df <- data.frame(xmin=0, xmax=0.15, ymin=-Inf, ymax=Inf)
  
secondplot <- mainplot + coord_cartesian(xlim=c(0, 0.15)) + 
    ggtitle("Lower Range of\nFractional Coverage of Agriculture\n(National Land Cover Database 2011")

mainplot <- mainplot + geom_rect(data = rect_df, color="khaki4", alpha=0.3,
                                 aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax)) +
  geom_text(data = data.frame(x1 = 0.25, y1 = 12, label = "See graph below\nfor a closer view"), 
            aes(x = x1, y = y1, label = label))

mainplot
secondplot

```


\newpage


```{r popdens, echo=FALSE}
```

##Population Density

Population density is based on the 2010 US Population Census and is measured in people per square kilometer. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

df_pop_AC <- format.df("landcover.AC_POPD10", "POPD10_AC", metabData, allData_pop, type=NULL)
df_pop_RE <- format.df("landcover.RE_POPD10", "MEAN", metabData, allData_pop, type=NULL)

df_pop <- format.catchment.df(df_pop_AC, df_pop_RE)
plot.dens(df_pop, "Population Density (2010 US Population Census)", log=TRUE, reach=TRUE, type="",
          xlabel="Population Density\n(people per square kilometer)")

```


\newpage


```{r eflow, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, eval=FALSE}

##Hydrologic Indicator Statistics

# A collection of hydrologic indicator statistics calculated using streamflow data for the 291 metabolism sites. The R package EflowStats, developed by Jessica Thompson and Stacey Archfield, was used to pull streamflow data from NWIS and calculate the stats. Some of the sites had incomplete data to calculate the statistics (see Table 2 below).   


#sites must be from the get_meta() because the csv dropped the leading 0

library(EflowStats)

not_working <- c("0166818623", "0166818985", "02110400", "02148315", "02162035",
                 "02336152", "02344630", "03259813", "04119400", "05406469", 
                 "05406479", "08437710", "08447300", "11463682", "11463980", 
                 "11465240", "13173600", "13206400")
working <- which(!metabData$site_num %in% not_working)

#eflow_metab <- eflow.mag7(metabData$site_num[working])
eflow_metab <- read.csv("data(DO_NOT_COMMIT)/eflow_metab.csv")

missing_table <- data.frame(`Site Number`=not_working)

var_table <- data.frame(Variable=c("lam1Obs", "tau2Obs", "tau3Obs", "tau4Obs", 
                                  "ar1Obs", "amplitudeObs", "phaseObs"),
                       Definition=c("Arithmetic mean", 
                                    "Coefficient of L-variation - analagous to coefficient of variation.",
                                    "The third L-moment ratio or L-skew.",
                                    "The fourth L-moment ratio or L-kurtosis.",
                                    "AR1 correlation coefficient.",
                                    "Amplitude of the seasonal signal. Compute seasonality variables by first
                                    standardizing flows, using the fitting relation A*cos(2*pi*t) + B*sin(2*pi*t).",
                                    "Phase of the seasonal signal. Compute seasonality variables by first 
                                    standardizing flows, the fitting relation A*cos(2*pi*t) + B*sin(2*pi*t)."))

```

```{r eflow_plots, echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE}
eflowPlot(eflow_metab, "")

kable(var_table, caption="Table 1. EflowStats variable definitions")

kable(missing_table, caption="Table 2. Metabolism sites not included in hydrologic indicator stats calculations.")

```

##Ecoregion

Below is a map showing the geographic location of the metabolism sites (red dots) and a count of the metabolism sites located in each of the 84 ecoregions of the contiguous US (color of ecoregion based on number of sites).  

```{r echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, results='hide'}


library(rgeos)
library(rgdal)
library(mda.streams)
library(dplyr)
library(unitted)
library(RColorBrewer)
library(httr)

plot.CRS <- "+init=epsg:2163"

missing.data = "#B3B3B3"
site.color = rgb(1,0,0,0.3)


# level III Ecoregions come from GDP geoserver

## -- get spatial data --
# ecoregions:
destination = tempfile(pattern = 'eco_shape', fileext='.zip')
query <- 'http://cida.usgs.gov/gdp/geoserver/wfs?service=WFS&request=GetFeature&typeName=derivative:Level_III_Ecoregions&outputFormat=shape-zip&version=1.0.0'
file <- GET(query, write_disk(destination, overwrite=T), progress())
shp.path <- tempdir()
unzip(destination, exdir = shp.path)
eco = readOGR(shp.path, layer='Level_III_Ecoregions') %>% 
  spTransform(CRS(plot.CRS))

# sites: 
metabDataAll <- get_meta()
metab.sites <- metabDataAll[which(metabDataAll$manual.assessment %in% c("accept", "examine")), ]$site_name

site.loc <- get_site_coords(metab.sites, format="normal")
site.points <- data.frame(lon=site.loc$lon, lat = site.loc$lat)
sp2 <- SpatialPoints(site.points, proj4string=CRS("+proj=longlat + datum=wgs84")) %>% 
  spTransform(CRS(plot.CRS))
## -- get spatial data /--

```

```{r echo=FALSE, eval=TRUE}
## -- analyze point in polygon --
point.in = gContains(eco, sp2, byid=TRUE)

counts.by.id = colSums(point.in)
## -- analyze point in polygon /--


## -- color markers -- 
bins = pretty(counts.by.id, 100)
key.bins = pretty(counts.by.id, 5)
pal = colorRampPalette(brewer.pal(9, 'YlGnBu'))(length(bins))
key.cols = colorRampPalette(brewer.pal(9, 'YlGnBu'))(length(key.bins))
```

```{r echo=FALSE, eval=TRUE}
if (bins[1] != 0 | key.cols[1] != 0)
  #stop('cannot assign zero color when there is no zero value')
pal[1] <- missing.data # 0 is grey
key.cols[1] <- missing.data  # 0 is grey
#get closest bin
bin = unname(sapply(counts.by.id, function(x) ifelse(is.na(x),NA,which.min(abs(x-bins)))))
cols = rep(NA, length(counts.by.id))
cols[!is.na(bin)] = pal[bin[!is.na(bin)]]
## -- color markers /-- 

```

```{r ecoregion_plots, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, fig.height=5.5, fig.width=7.2}

library(DT)
layout(matrix(data = c(1,1,1,1,1,1,2), ncol=1))
par(mai = c(0,0,0,0), omi = c(0,0,0,0))

xlim <- c(-1534607.9,2050000.1) # specific to the transform we are using
ylim <- c(-2072574.6,727758.7)

plot(eco, add = FALSE, col = cols, border = 'grey35', lwd = 0.5, xlim = xlim, ylim = ylim)
plot(sp2, add = TRUE, col=site.color, pch=20, cex=0.5)


# secondary plot for color legend
plot(c(NA,NA),c(NA,NA), axes=F, ylim=c(0,1),xlim=c(0,1))
bin.w = 0.07
spc = .02
text(.1,.5, 'Number of sites', pos=3, offset=0.1)
for(i in 1:length(key.cols)){
  x1 = 0.20+(i-1)*(bin.w+spc)
  graphics::rect(x1, 0.3, x1+bin.w, 0.8, col=key.cols[i], lwd=NA)
  text(x1+bin.w/2, y=0.33, labels=key.bins[i], pos=1)
}
## -- plot map /-- 
eco.table = data.frame('Ecoregion'=as.character(eco$LEVEL3_NAM), 'Number of sites'=counts.by.id)
row.names(eco.table) <- NULL
srt.i = sort(eco.table[, 2], decreasing = TRUE, index.return = TRUE)$ix
DT::datatable(eco.table[srt.i, ], rownames = FALSE)

```



##State Discharge & DO 

The following state maps depict the geographic locations for dissolved oxygen (DO) and discharge sites. The states shown below are New Hampshire, Puerto Rico, North Carolina, Florida, Wisconsin, Arizona, and Connecticut.   

```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}


library(leaflet)
library(dataRetrieval)
library(DT)

disch.code = unique(get_var_src_codes(var=="disch", out='p_code'))
doobs.code = "00300"
min.count = 100
skip.types = c('GW','LK','ES')

get_site <- function(stateCd, pCd){
  filter(readNWISdata(service = "site", stateCd=stateCd,seriesCatalogOutput="true",outputDataTypeCd="iv",parameterCd=pCd,Access="3"), parm_cd==pCd) %>% 
    filter(count_nu > min.count) %>% 
    filter(!(site_tp_cd %in% skip.types)) %>% 
    filter(!is.na(dec_long_va))
}

make_map <- function(state){
  state.cd = stateCd$STUSAB[stateCd$STATE_NAME == state]
  
  
  clean_site <- function(site){
    clean.site= select(site, site_no, station_nm, dec_lat_va, dec_long_va, loc_web_ds, begin_date, end_date, count_nu) %>%
      rename(days_recorded = count_nu, device=loc_web_ds, latitude=dec_lat_va, longitude=dec_long_va) %>% 
      mutate(name=gsub("\\b([a-z])([a-z]+)", "\\U\\1\\L\\2" ,tolower(station_nm), perl=TRUE)) %>% 
      mutate(site_no = paste0('nwis_',site_no)) %>% 
      select(-station_nm)
    return(clean.site)
  }
  
  do.sites = get_site(state.cd, doobs.code) %>% 
    clean_site()
  disch.sites = get_site(state.cd, disch.code) %>% 
    clean_site()
  
  m = leaflet() %>% 
    addProviderTiles("CartoDB.Positron",
                       options = leaflet::providerTileOptions(noWrap = TRUE)) %>% 
    addCircleMarkers(lng = do.sites$longitude, 
                     lat = do.sites$latitude, 
                     popup = do.sites$station_nm, color = '#ff0000') %>% 
    addCircleMarkers(lng = disch.sites$longitude, 
                     lat = disch.sites$latitude, 
                     popup = disch.sites$station_nm, color = 'blue', radius = 3) %>% 
    addLegend("bottomright", labels = c('DO', 'Discharge'), 
              colors = c('red', 'blue'), title = 'Metabolism Site Types')
  
  return(list(map=m, disch=disch.sites, do = do.sites))
}
```

####New Hampshire
```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}
mapper = make_map('New Hampshire')
print(mapper$map)
```
DO stations
```{r}
datatable(mapper$do, rownames = FALSE)
```
Discharge stations
```{r}
datatable(mapper$disch, rownames = FALSE)
```

####Puerto Rico
```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}
mapper = make_map('Puerto Rico')
print(mapper$map)
```
DO stations
```{r}
datatable(mapper$do, rownames = FALSE)
```
Discharge stations
```{r}
datatable(mapper$disch, rownames = FALSE)
```


####North Carolina
```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}
mapper = make_map('North Carolina')print(mapper$map)
```
DO stations
```{r}
datatable(mapper$do, rownames = FALSE)
```
Discharge stations
```{r}
datatable(mapper$disch, rownames = FALSE)
```


####Florida
```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}
mapper = make_map('Florida')
print(mapper$map)
```
DO stations
```{r}
datatable(mapper$do, rownames = FALSE)
```
Discharge stations
```{r}
datatable(mapper$disch, rownames = FALSE)
```


####Wisconsin
```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}
mapper = make_map('Wisconsin')print(mapper$map)
```
DO stations
```{r}
datatable(mapper$do, rownames = FALSE)
```
Discharge stations
```{r}
datatable(mapper$disch, rownames = FALSE)
```


####Arizona
```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}
mapper = make_map('Arizona')
print(mapper$map)
```
DO stations
```{r}
datatable(mapper$do, rownames = FALSE)
```
Discharge stations
```{r}
datatable(mapper$disch, rownames = FALSE)
```


####Connecticut
```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}
mapper = make_map('Connecticut')
print(mapper$map)
```
DO stations
```{r}
datatable(mapper$do, rownames = FALSE)
```
Discharge stations
```{r}
datatable(mapper$disch, rownames = FALSE)
```

